---
title: "NYCflights13 Exercises"
author: "Sebnem Er"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("~/Desktop/MSc Data Science/EDA/flights.rda")
```

## NYCFlights13 Dataset

### flights

Description

On-time data for all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013.

Data frame with columns

* year, month, day Date of departure.
* dep_time, arr_time Actual departure and arrival times (format HHMM or HMM), local tz.
* sched_dep_time, sched_arr_time Scheduled departure and arrival times (format HHMM or HMM),
local tz.
* dep_delay, arr_delay Departure and arrival delays, in minutes. Negative times represent early
departures/arrivals.
* carrier Two letter carrier abbreviation. See airlines to get name.
* flight Flight number.
* tailnum Plane tail number. See planes for additional metadata.
* origin, dest Origin and destination. See airports for additional metadat
* air_time Amount of time spent in the air, in minutes.
* distance Distance between airports, in miles.
* hour, minute Time of scheduled departure broken into hour and minutes.
* time_hour Scheduled date and hour of the flight as a POSIXct date. Along with origin, can be
used to join flights data to weather data.

https://cran.r-project.org/web/packages/nycflights13/nycflights13.pdf
```{r summary}
library(dplyr)
summary(flights)
head(flights, n=10)

```
## Exercises

1. Find all flights that

a) Had an arrival delay of two or more hours

```{r}

arr_delay_2hrs <- flights %>% filter(arr_delay > 120)
head(arr_delay_2hrs)

```

b) Flew to Houston (IAH or HOU)

```{r}

houston_dest <- flights %>% filter(dest == 'IAH' | dest == 'HOU')
head(houston_dest)
unique(houston_dest$dest)
```

c) Were operated by United, American, or Delta
```{r}
UAD_airlines <- flights %>% filter(carrier == 'DL' | carrier == 'UA' | carrier == 'AA' )
head(UAD_airlines)
unique(UAD_airlines$carrier)
```
d) Departed in summer (July, August, and September)
```{r}
summer_flights <- flights %>% filter(month %in% c(7, 8, 9))
head(summer_flights)
unique(summer_flights$month)
```
e) Arrived more than two hours late, but didn’t leave late
```{r}
arr_late_dep_ontime <- flights %>% filter(arr_delay > 120 & dep_delay <= 0 )
arr_late_dep_ontime
```
f) Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
flights_diff <- flights %>% mutate(diff = (dep_delay - arr_delay))
flights_30diff <- flights_diff %>% filter(dep_delay > 60 & diff >= 30)
head(flights_30diff)

```
g) Departed between midnight and 6am (inclusive)
```{r}
morning_flights <- flights %>% filter(dep_time %in% seq(from=0, to=600))
head(morning_flights)

```

2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
```{r}
morning_flights_between <- flights %>% filter(between(dep_time,0,600))
head(morning_flights_between)
```

3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
```{r}
missing_dep_time <- flights[is.na(flights$dep_time),]
nrow(missing_dep_time)
```
#### Other missing variables include dep_delay, arr_time,arr_delay.

4. Sort the flights dataframe according to day, month and year
```{r}
flights_sort_date <- flights %>% arrange(day,month,year)
head(flights_sort_date)
```

5. Sort the flights dataframe using the arrival time in a descending order.
```{r}
flights_sort_arr_time <- flights %>% arrange(desc(arr_time))
head(flights_sort_arr_time)
```

6. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
flights_na_sort <- flights %>% arrange(!is.na(arr_time),arr_time)
head(flights_na_sort)

```

7. Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
flights_delayed <- flights %>% arrange(desc(arr_delay)) %>% filter(arr_delay >= 0)
flights_delayed

flights_left_early <- flights %>% arrange(arr_delay) %>% filter(arr_delay <= 0)
head(flights_left_early)
```

8. Sort flights to find the fastest (highest speed) flights.
```{r}
flights_speed <- flights %>%  mutate(time_taken = (hour*60)+ minute) %>% mutate(speed = (distance/time_taken)) %>% arrange(desc(speed))
head(flights_speed)
```

9. Which flights travelled the farthest? Which travelled the shortest?
```{r}

```

10. Select all columns in the flights dataframe between year and day (inclusive).
```{r}

```

11. Select all columns except those from year to day (inclusive).
```{r}

```

12. Rename the tail_num variable in flights dataframe with tailnum.
```{r}

```

13. Using the pipeline operator do the following:

a) Select all columns in the flights dataframe between year and day (inclusive).
```{r}

```
b) Select all columns that ends with delay and time.
```{r}

```
c) Select the distance and air_time variables.
```{r}

```
d) Think about a way of creating a gain/loss travel time for each flight and create this variable as ``gain".
```{r}

```
e) What is the speed of the flight.
```{r}

```

14. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}

```

15. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?
```{r}

```

16. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?
```{r}

```

17. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().
```{r}

```

18. What does 1:3 + 1:10 return? Why?
```{r}

```

19. Group the flights with year, month and day and then summarise the groupings with count, mean distance and mean arr_delay. Filter to remove noisy points and Honolulu airport. Filter the flights that took place more than 20 times. Plot distance vs delay using ggplot.
```{r}

```

20. Group flights by destination. Summarise to compute distance, average delay, and number of flights.
```{r}

```

21. Which plane (tailnum) has the worst on-time record?
```{r}

```

22. What time of day should you fly if you want to avoid delays as much as possible?
```{r}

```

23. For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.
```{r}

```

24. Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using lag(), explore how the delay of a flight is related to the delay of the immediately preceding flight.
```{r}

```

25. Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time of a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?
```{r}

```

26. Find all destinations that are flown by at least two carriers. Use that information to rank the carriers.
```{r}

```

27. For each plane, count the number of flights before the first delay of greater than 1 hour.

```{r}

```
