# EDA Lecture 1-2 R Examples

## Example 2 - Gapminder

```{r}
#install.packages("gapminder")
library(gapminder)
gapminder
```

\pagebreak

## Frequency Distribution
```{r}
library(ggplot2)
table(gapminder$continent)
```


## Bar Plot

```{r}
library(ggplot2)
plot1 <- ggplot(gapminder, aes(x=continent)) + geom_bar()
plot1
```

## Pie Chart

```{r}
plot1 + coord_polar()
```

If you would like to have a regular pie chart, then you need to provide the frequency distribution.

## Histogram
### A Simple Histogram

```{r}
library(ggplot2)
plot2 <- ggplot(gapminder,
                aes(x = gdpPercap))
plot2 + geom_histogram(binwidth = 1000)
```

\pagebreak

### Histogram With a Title

```{r}
plot2 + 
  geom_histogram(binwidth = 1000) +
  labs(title = "Histogram of GDP per Capita for All Years")
```


\pagebreak

### Histogram with Different Color Schemes:

```{r}
plot2 + 
  geom_histogram(binwidth = 1000, color="black", fill="white") +
  labs(title = "Histogram of GDP per Capita for All Years")
```



\pagebreak

### Histogram of Log Transformed Variable:

```{r}
plot3 <- ggplot(gapminder,
                aes(x = log10(gdpPercap)))
plot3 + 
  geom_histogram(binwidth = .2, color="black", fill="white") +
  labs(title = "Histogram of  Log Transformed GDP per Capita for All Years")
```

### Determine the Binwidth

How do we determine the binwidth?
\begin{itemize}
\item Sturges' rule uses class intervals of length \\
$L = \frac{x_{max}-x_{min}}{1+1.44*ln(n)}$ \\
\item Genstat rule uses uses class intervals of length: \\
$L = \frac{x_{max}-x_{min}}{\sqrt{n}}$ \\
\item or a general rule
\end{itemize}

So we can create our own function for the binwidth:

```{r}
width_bin = function(x) (max(x)-min(x)) / sqrt(length(x))
manualbin = width_bin(log10(gapminder$gdpPercap))
```

\pagebreak
```{r}
plot3 + 
  geom_histogram(binwidth = manualbin, color="black", fill="white") 
```

\pagebreak

or simply

```{r}
plot3 + 
  geom_histogram(binwidth = function(x) (max(x)-min(x)) / sqrt(length(x)), color="black", fill="white") +
    labs(title = "Histogram of  Log Transformed GDP per Capita for All Years (Genstat Binwidth)")

```

But you will notice that Gdp per capita variable includes all years, all continents, all countries!!!

\pagebreak

### Histogram for a Subset of Data

Log Transformed GDP per Capita for 2007:

```{r}
plot4 <- ggplot(subset(gapminder, year == 2007),
                aes(x = log10(gdpPercap)))
plot4 + 
   geom_histogram(binwidth = function(x) (max(x)-min(x)) / sqrt(length(x)), color="black", fill="white") +
  labs(title = "Histogram of  Log Transformed GDP per Capita for 2007 (Genstat Binwidth)")
```


\pagebreak

### Histogram with Overall Mean Line 

Log Transformed GDP per Capita for 2007 with the Overall Mean Line
 
```{r}
# Histogram with mean of log10(gdpPercap) on the plot
plot4 + 
  geom_histogram(binwidth = function(x) (max(x)-min(x)) / sqrt(length(x)), color="black", fill="white") +
  geom_vline(aes(xintercept=mean(log10(gdpPercap))),
            color="blue", linetype="dashed", size=1) +
  labs(title = "Histogram of  Log Transformed GDP per Capita for 2007 (Genstat Binwidth)")

```

\pagebreak

### Histogram with Density plot

```{r}
# Histogram with density plot
ggplot(subset(gapminder, year == 2007),
                aes(x = log10(gdpPercap))) +  
  geom_histogram(aes(y=..density..), binwidth = function(x) (max(x)-min(x)) / sqrt(length(x)),colour="black", fill="white")+
  geom_density(alpha=0, fill="#FF6666") #alpha for transparency, if alpha = 0, no fill
```

\pagebreak

### Histogram with Facets

How about looking at the differences among different continents?

```{r}
# Histogram with mean of log10(gdpPercap) on the plot
plot4 + 
  geom_histogram(binwidth = function(x) (max(x)-min(x)) / sqrt(length(x)), color="black", fill="white") +
  geom_vline(aes(xintercept=mean(log10(gdpPercap))),
            color="blue", linetype="dashed", size=1) +
  facet_grid(continent ~ .)
```

\pagebreak

## Boxplots

```{r}
# Histogram with mean of log10(gdpPercap) on the plot
plot5 <- ggplot(subset(gapminder, year == 2007),
                aes(x = year, y = log10(gdpPercap))) 
# if x axis variable is numeric, then one single boxplot
# if x axis variable is categorical, then works like facets

plot5 + 
  geom_boxplot() #+ coord_flip()
```

Try with ``continent" variable.

## Scatter Plots

### A Simple Scatter Plot

```{r}
plot6 <- ggplot(subset(gapminder, year == 2007),
                aes(x = lifeExp, y = log10(gdpPercap)))
plot6 +
  geom_point()
```

### Scatter Plot with Labellings

```{r}
plot6 +
  geom_point(aes(colour = factor(continent)))
```

### Scatter Plot with Linear Lines for Different Groups

```{r}
plot6 +
  geom_point(aes(colour = factor(continent))) +
  geom_smooth(aes(group = continent, colour = factor(continent)), lwd = 1, se = FALSE, method = "lm")
```

```{r}
plot7 <- ggplot(gapminder,
                aes(x = lifeExp, y = log10(gdpPercap)))
plot7 +
  geom_point(aes(colour = factor(continent))) +
  facet_wrap(~ year) # scales = "free_x"
```




For more check ``ggplot2: Elegant Graphics for Data Analysis (Use R!)" by (Hadley Wickham)
